\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{indentfirst}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{float}
\usepackage[a4paper, margin = 2 cm]{geometry}
\usepackage{fancyhdr}
\usepackage{wrapfig}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algpseudocode}

\title{Languages, Automata and Computation II assignment 3}
\author{Dominik Wawszczak}
\date{2025-01-31}

\begin{document}
	\setlength{\parindent}{0 cm}
	
	Dominik Wawszczak \hfill Languages, Automata and Computation II
	
	Student ID Number: 440014 \hfill Assignment 3
	
	Group Number: 1
	
	\bigskip
	\hrule
	\bigskip
	
	\textbf{Problem 1}
	
	\medskip
	
	This problem was solved in collaboration with Kacper Bal and Mateusz
	Mroczka.
	
	\medskip
	
	We say that a language \(L \in \mathbb{A}^{*}\) is a \textit{good}, if it
	satisfies the condition from the statement, i.e.
	\[ \underset{w \in \mathbb{A}^{\ast}}{\forall} \
	\underset{\sigma : \mathbb{A} \to \mathbb{A}}{\forall} \ (w \in L \ \iff \
	\sigma(w) \in L) \text{.} \]
	
	\medskip
	
	Let \(A \in \mathbb{A}\) be an arbitrary element. Define \(\sigma_{A}\) as
	the function constantly equal to \(A\). For any good language \(L \in
	\mathbb{A}^{\ast}\), it must hold that
	\[ \underset{w \in \mathbb{A}^{\ast}}{\forall} \ (w \in L \ \iff \
	\sigma_{A}(w) \in L) \text{,} \]
	since the order of quantifiers does not matter. This is equivalent to
	\[ \underset{w \in \mathbb{A}^{\ast}}{\forall} \ \big( w \in L \ \iff \
	A^{|w|} \in L \big) \text{,} \]
	which implies that for every \(n \in \mathbb{N}\), \(L\) either contains all
	words of length \(n\) or none of them.
	
	\medskip
	
	The problem of determining whether a language \(L \in \mathbb{A}^{\ast}\) is
	good is semi-decidable, since we can iterate through register automata that
	use zero registers, comparing each one with the automaton from the input.
	
	\medskip
	
	The complement of this problem is also semi-decidable, as we can iterate
	over words of length \(n\), for \(n \in \mathbb{N}\), considering only a
	subset of \(n\) symbols from \(\mathbb{A}\). For each such word, we can
	check all the possible functions \(\sigma\) limited to this subset and
	verify whether the condition holds.
	
	\medskip
	
	Since both the problem and its complement are semi-decidable, the problem is
	decidable, which completes the proof.
	
	\bigskip
	
	\textbf{Problem 2}
	
	\medskip
	
	This problem was solved in collaboration with Kacper Bal and Mateusz
	Mroczka.
	
	\begin{enumerate}
		\item Let \(L = \{b^{\ast} a\}^{\omega}\), and \(w = b^{\omega}\).
		      Clearly, \(w \notin L\), but any finite prefix of \(w\) can be
		      extended to a word in \(L\) by concatenating it with
		      \(a^{\omega}\). Therefore, \(L\) is \(\omega\)-regular, but not
		      closed, which concludes the proof.
		
		\item Let \(w = a^{1} b^{1} a^{2} b^{2} a^{3} b^{3} \ldots\), and \(L =
		      \{w\}\). As shown in the tutorial, \(L\) is not
		      \(\omega\)-regular. To prove that it is closed, consider any \(v
		      \in \Sigma^{\omega}\) different from \(w\). Let \(n \in
		      \mathbb{N}\) be the smallest index at which \(w\) and \(v\)
		      differ. Then, the prefix of \(v\) of length \(n + 1\) (assuming
		      \(0\)-indexing) cannot be extended to an \(\omega\)-word in \(L\).
		      Additionally, any finite prefix of \(w\) can be extended to \(w\).
		      Hence, \(L\) is closed, which completes the proof.
		
		\item We can construct a parity automaton that is equivalent to the
		      Büchi automaton from the input, as discussed in the lecture. Let
		      \(\mathcal{A}\) be this parity automaton, and let \(A\) be the set
		      of states from which there exists an accepting run. For each state
		      \(q\), we can verify whether \(q \in A\) by setting the initial
		      state to \(q\) and checking for nonemptiness.
		      
		      Let \(\mathcal{A}'\) be the automaton obtained from
		      \(\mathcal{A}\) by keeping only the states in \(A\), and
		      increasing every weight by one. By checking the nonemptiness of
		      \(\mathcal{A}'\), we can determine whether there exists a run in
		      \(\mathcal{A}\) that visits only the states from \(A\) but is not
		      accepting. If such a run exists, the language \(L(\mathcal{A})\)
		      is not closed, as there exists an \(\omega\)-word corresponding to
		      this run, whose every finite prefix can be extended to an
		      \(\omega\)-word in \(L(\mathcal{A})\), but which is not in the
		      language itself. If no such run exists, the language is closed,
		      which concludes the proof.
	\end{enumerate}
\end{document}
